A flexible Prolog interpreter in Python

### 2007-05-01 talk/badhonnef2007/talk.tex
reworking the talk

  \begin{itemize}
  \item
    Pyrolog is a Prolog interpreter written in RPython
  \item
    RPython is a subset of Python translatable to other languages
  \item
    translation part done with the help of the PyPy project
  \end{itemize}
\end{frame}

\section{The PyPy Approach to VM Construction}
\subsection{Overview}
\begin{frame}
  \frametitle{What is PyPy?}
  \begin{itemize}
  \item
    started as a Python VM implementation
    in RPython (a well-chosen subset of Python)
  \item
    includes a translation tool-chain
  \item
    is becoming a general environment for writing interpreters (JavaScript, Prolog started)
  \item
    Open source project (MIT license)
  \item
    received EU funding for 2.5 years
  \end{itemize}

@@

\subsection{Motivation}
\begin{frame}
  \frametitle{VMs are still hard}
  It is hard to achieve:

@@

  \begin{itemize}
  \item
    Pervasive decisions: reference counting, single global lock ...
  \item
    No dynamic compilation
%  \pause
  \item
    Extensions:
    \begin{itemize}
    \item
      \alert{Stackless} (unlimited recursion, coroutines, serializable continuations)
    \item
      \alert{Psyco} (run-time specializer)
    \item
      \alert{Jython}, \alert{IronPython}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Python case (ii)}
  \begin{itemize}
  \item
    Extensions have problems
    \begin{itemize}
    \item
      need to keep track of CPython
    \item
      are hard to maintain
    \item
      Psyco very hard to port to other hardware architectures
    \end{itemize}
  \item
    The community wants Python to run everywhere:
    Jython (Java), IronPython (.NET).
    Lots of effort and duplication.

@@

\begin{frame}
  \frametitle{The Prolog case}
  \begin{itemize}
  \item
    problem mitigated by the fact that Prolog the language does not change
  \item
    a lot of implementations out there
  \item
    well-tuned mature C implementations (Sictsus, XSB, SWI, GNU-Prolog)
    \begin{itemize}
    \item
      have sometimes incompatible extensions to core Prolog
    \item
      interfacing with libraries is tedious
    \item
      changing the language to experiment is hard
    \item
      fixed implementation decisions (GC, how to generate code, etc.)
    \end{itemize}
  \item
    on CLR (P\#) and JVM (Prolog Café, tuProlog)
    \begin{itemize}
    \item
      interfacing with libraries of the platform mostly easy
    \item
      no extensions to core Prolog (like tabling, coroutines)
    \item
      slow, compared to good C implementations
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Mapping to low-level targets}
  \begin{itemize}
  \item
    Mechanically translate the interpreter to multiple
    lower-level targets
    \begin{itemize}
    \item C-like
    \item Java
    \item .NET
    \end{itemize}
  \item
    Insert low-level aspects into the code as required by
    the target (Object layout, memory management)
    \begin{itemize}
    \item object layout
    \item memory management
    \end{itemize}  \item
    Optionally insert new pervasive features not expressed
    in the source
    \begin{itemize}
    \item continuations
    \item dynamic compilation
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Steps}
  \begin{columns}[c]
  \begin{column}{5cm}
    \includegraphics[width=5cm]{image/arch.pdf}
  \end{column}
  \begin{column}{7cm}
  \begin{itemize}
    \item
      Generate flow graphs from the RPython program
    \item
      Peform global type inference on the flow graphs
    \item
      Transform flow graphs through several steps until they match the level of
      the target environment
    \item
      Weave in translation aspects in the process
  \end{itemize}
  \end{column}
  \end{columns}

\begin{frame}
  \frametitle{Translation Aspects (i)}
  Features not present in the source can be
  added during translation.

  Example: memory management:
  \begin{itemize}
  \item
  Boehm garbage collector
  \item
  mark-n-sweep written in RPython, with additional features
  \item
  reference counting
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Aspects (ii)}
  \begin{itemize}
  \item
    \alert{Stackless transformation}: continuation capture, implemented by
    saving the low-level frames' local variables into the heap and back
    \begin{itemize}
      \item allows arbitrarily deep stack usage
      \item uses the C stack as long as possible
      \item has the consequence of making RPython do tail call elimination
    \end{itemize}
  \item
    work in progress: turning an interpreter into a just-in-time compiler
    is a translation aspect too
  \end{itemize}
\end{frame}

\section{The Prolog interpreter}
\begin{frame}
  \frametitle{Prolog Interpreter Implementation}
  \begin{itemize}
  \item
    naive, very simple interpreter
    \begin{itemize}
    \item uses "structure copying"
    \item interprets Prolog terms directly, no bytecode
    \end{itemize}
  \item
    uses continuation passing style similar to BinProlog
  \item
    Prolog calls mapped to RPython calls
    \begin{itemize}
    \item possible because stackless allows arbitrary deep recursion
    \end{itemize}
  \item
    implements large parts of the ISO standard (some builtins missing)
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Interpreter Facts}
  \begin{itemize}
  \item
    2500 lines of Python code in total
  \item
    700 of those are for builtins
  \item
    after translation to C: 14000 line of C code
  \item
    part of the PyPy distribution at:
    http://codespeak.net/pypy
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Performance (i)}
  \includegraphics[scale=0.4]{image/bench.png}
\end{frame}

\begin{frame}
  \frametitle{Performance (ii)}
  \begin{itemize}
  \item
    performance is quite bad compared to tuned C implementations
  \item
    performance is pretty good compared to Java and .NET implementations
  \item
    surprising, since those are often based on the WAM
  \item
    maybe it's hard to simulate the WAM on such a VM
  \end{itemize}
\end{frame}

@@

\section*{Summary}
\subsection*{Summary}
\begin{frame}
  \frametitle<presentation>{Summary}

  % Keep the summary *very short*.
  \begin{itemize}
  \item
    The construction of virtual machines gets easier when using high-level
    languages
  \item
    XXX
  \item
    XXX
  \end{itemize}
\end{frame}
\subsection*{Outlook}
\begin{frame}
  \frametitle<presentation>{Outlook}
  % The following outlook is optional.
  \begin{itemize}
  \item
  \end{itemize}
\end{frame}

>>>>

  \begin{itemize}
  \item
    Pyrolog is a Prolog interpreter written in RPython
  \item
    RPython (``Restricted Python'') is a subset of Python translatable to other
    languages
  \item
    RPython is designed to be significantly faster than regular Python
  \item
    translation tool-chain part of the PyPy project
  \end{itemize}
\end{frame}

\section{The PyPy Approach to VM Construction}
\subsection{Overview}
\begin{frame}
  \frametitle{What is PyPy?}
  \begin{itemize}
  \item
    started as a Python VM implementation in Python
  \item
    Python itself is too dynamic to be translatable to other languages, need a subset
  \item
    includes a translation tool-chain for RPython
  \item
    is becoming a general environment for writing interpreters (JavaScript, Prolog started)
  \item
    Open Source project (MIT license)
  \item
    received EU funding for 2.5 years
  \end{itemize}

@@

\subsection{Motivation}
\begin{frame}
  \frametitle{VMs are still hard}
  Hard to reconcile:

@@

  \begin{itemize}
  \item
    Pervasive decisions: reference counting, global lock \dots
  \item
    No dynamic compilation
%  \pause
  \end{itemize}
  \begin{block}{
    Extensions:}
    \begin{itemize}
    \item
      \alert{Stackless} (unlimited recursion, coroutines, serializable continuations)
    \item
      \alert{Psyco} (run-time specializer)
    \item
      \alert{Jython}, \alert{IronPython}
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{The Python case (ii)}
  \begin{block}{Problems of Extensions:}
    \begin{itemize}
    \item hard to maintain: need to keep track of CPython
    \item Psyco very hard to port to other hardware architectures
    \item tedious to write them, Python semantics need to be re-implemented
    \end{itemize}
  \end{block}
  \begin{itemize}
  \item
    The community wants Python to run everywhere:
    Jython (Java), IronPython (.NET).
    Lots of effort and duplication.

@@

\begin{frame}
  \frametitle{The Prolog case (i)}
  \begin{itemize}
  \item
    problem mitigated by the fact that Prolog the language does not change
  \item
    the core of Prolog is very simple (at least compared to Python)
  \item
    a lot of implementations out there
  \item
    well-tuned mature C implementations (Sicstus, XSB, SWI, GNU-Prolog)
  \item
    on CLR (P\#) and JVM (Prolog Café, tuProlog)
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Prolog case (ii)}
  \begin{block}{mature C implementations}
    \begin{itemize}
    \item
      interfacing with libraries is tedious
    \item
      changing the language to experiment is hard
    \item
      often extensions to core Prolog, incompatible between each other
    \item
      fixed implementation decisions (GC, how to generate code, etc.)
    \end{itemize}
  \end{block}
  \begin{block}{implementations on CLR and JVM}
    \begin{itemize}
    \item
      interfacing with libraries of the platform mostly easy
    \item
      no extensions to core Prolog (like tabling, coroutines)
    \item
      slow, compared to good C implementations
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Mapping to low-level targets}
  \begin{itemize}
  \item
    Mechanically translate the interpreter to multiple
    lower-level targets
    \begin{itemize}
    \item C-like
    \item Java
    \item .NET
    \end{itemize}
  \item
    Insert low-level aspects into the code as required by
    the target (Object layout, memory management)
    \begin{itemize}
    \item object layout
    \item memory management
    \end{itemize}  \item
    Optionally insert new pervasive features not expressed
    in the source
    \begin{itemize}
    \item continuations, ``micro-threads''
    \item dynamic compilation
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Aspects (i)}
  Features not present in the source can be added during translation.
  \begin{itemize}
  \item
    \alert{memory management}: use different GC strategies (Boehm collector, custom mark-n-sweep)
  \item
    \alert{Stackless transformation}: allows program to control its stack (continuations, \dots)
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Translation Aspects (ii)}
  A \alert{JIT compiler} as a translation aspect
  \begin{itemize}
  \item
    Transform the interpreter into a JIT compiler, using partial evaluation and
    specialization techniques
  \item
    Some hints in the interpreter source needed
  \item
    Current prototype applied to Python interpreter gives impressive speedups
  \end{itemize}
\end{frame}

\section{The Prolog interpreter}
\subsection{Interpreter}
\begin{frame}
  \frametitle{Prolog Interpreter Implementation}
  \begin{itemize}
  \item
    naive, very simple interpreter
    \begin{itemize}
    \item uses "structure copying"
    \item interprets Prolog terms directly, no bytecode
    \end{itemize}
  \item
    uses continuation passing style inspired by BinProlog
  \item
    Prolog calls mapped to RPython calls
  \item
    implements large parts of the ISO standard (some builtins missing)
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Interpreter Facts}
  \begin{itemize}
  \item
    2500 lines of Python code in total
  \item
    700 of those are for builtins
  \item
    after translation to C: 14000 line of C code
  \item
    part of the PyPy distribution at:
    \texttt{http://codespeak.net/pypy}
  \end{itemize}
\end{frame}

\subsection{Performance}
\begin{frame}
  \frametitle{Performance (i)}
  \includegraphics[scale=0.4]{image/bench.png}
\end{frame}

\begin{frame}
  \frametitle{Performance (ii)}
  \begin{itemize}
  \item
    performance is quite bad compared to tuned C implementations
  \item
    performance is pretty good compared to Java and .NET implementations
  \item
    surprising, since those are often based on the WAM
  \item
    maybe the WAM model does not match these VMs very well?
  \end{itemize}
\end{frame}

\section*{Summary}
\subsection*{Summary}
\begin{frame}
  \frametitle{Summary}

  % Keep the summary *very short*.
  \begin{itemize}
  \item
    Very simple Prolog interpreter in RPython can compete with interpreters on the JVM, CLR
  \item
    Interpreter implementation eased by use of a high-level language
  \item
    Low-level details abstracted away but re-inserted later
  \end{itemize}
\end{frame}

\subsection*{Outlook}
\begin{frame}
  \frametitle{Outlook}
  \begin{itemize}
  \item
    complete the set of builtins
  \item
    tight language integration between Prolog and Python
  \item
    apply the dynamic compiler generator to the Prolog interpreter
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Backup slides}
  \dots
\end{frame}

\begin{frame}
  \frametitle{Translation Steps}
  \begin{columns}[c]
  \begin{column}{5cm}
    \includegraphics[width=5cm]{image/arch.pdf}
  \end{column}
  \begin{column}{7cm}
  \begin{itemize}
    \item
      Generate flow graphs from the RPython program
    \item
      Peform global type inference on the flow graphs
    \item
      Transform flow graphs through several steps until they match the level of
      the target environment
    \item
      Weave in translation aspects in the process
  \end{itemize}
  \end{column}
  \end{columns}
\end{frame}

@@
